### 2021-10-14 paper/paper.tex
changes

\newcommand{\fixme}[1]{\noindent{\color{red}\textbf{FIXME:}  {#1}}}
\newcommand{\fixmemike}[1]{\noindent{\color{blue}\textbf{FIXME (Mike):}  {#1}}}

@@

\begin{abstract}
    %Multiclass logistic regression has generalization error $O(\sqrt{k/n})$, where $k$ is the number of classes and $n$ the number of data points.

    Multiclass classification problems often have many semantically similar classes.
    For example, 90 of ImageNet's 1000 classes are for different breeds of dog.
    We should expect that these semantically similar classes will have similar parameter vectors,
    but the standard cross entropy loss does not enforce this constraint.
    %These classes are semantically similar to each other,
    %and so we expect their learned parameter vectors to also be similar.
    %Unfortunately, the standard cross entropy loss function does not enforce this constraint.
    %It has generalization error $O(\sqrt{k/n})$, where $k$ is the number of classes and $n$ the number of data points.

@@

@@

\end{itemize}

@@

\section{The Tree Loss}

@@

\section{Theoretical Results}

@@

\begin{lemma}
    \ref{lemma:starV}
    We have that for all $V$-trees derived from a $U$-tree,
    \begin{equation}
        \lF{\star V} \le \lF{\star U}.
    \end{equation}
\end{lemma}
\begin{proof}
    \fixme{}
\end{proof}

@@

\begin{lemma}
    \ref{lemma:starU}
    Under assumption \ref{ass:B},
    the following bound holds for all $U$-tree structures:
    \begin{equation}
        \lF{\star U} \le 2\sqrt{k}B.
    \end{equation}
\end{lemma}
\begin{proof}
    We have
    \begin{align}
        \lF{\star U}^2 
        &= \sum_{i=1}^k \ltwo{\star\uu_i}^2 \\
        &= \sum_{i=1}^k \ltwo{\star\w_i - \star\w_{\parent{i}}}^2 \\
        &\le \sum_{i=1}^k \left(\ltwo{\star\w_i} + \ltwo{\star\w_{\parent{i}}} \right)^2 \\
        &\le \sum_{i=1}^k (2B)^2 \\
        &\le k (2B)^2
        .
    \end{align}
\end{proof}

@@

\begin{lemma}
    \ref{lemma:main}
    When $c\le1$, we have that
    \begin{equation}
        \lF{\star U} \le \tfrac{1}{\sqrt2}\lambda B \sqrt{\log_2 k}.
        \label{eq:c<=1}
    \end{equation}
    When $c>1$, we have that
    \begin{equation}
        \lF{\star U} \le \sqrt{5}\lambda B \sqrt{k^{(1-1/c)}}.
        \label{eq:c>1}
    \end{equation}
\end{lemma}
\begin{proof}
%Construct a cover tree over the distance metric $d$.
%For each class $i$, define the vector
%\begin{equation}
    %\uu_i =
    %\begin{cases}
        %\w_i & \text{if $i$ is the root of the cover tree} \\
        %\w_i - \w_{\parent{i}} & \text{otherwise} \\
    %\end{cases}
    %.
%\end{equation}
%Define the vector $\star\uu_i$ similarly by replacing all the $\w_i$ with $\star\w_i$ above.
    First, we review some simple properties of the cover tree structure that will be needed in our proof.
    Let $q$ denote the maximum number of children of any node in the cover tree.
    A standard property of cover trees is that $q \le (4^c)$ \citep[Lemma 25]{izbickithesis}.
    Let $Q_i$ denote the number of nodes at level $i$ in the cover tree.
    Then we have that $|Q_i| \le q^i.$
    Finally, let $r$ be the height of the cover tree.
    Without loss of generality, we analyze the bound when the cover tree is perfectly balanced and so $r = \log_q k = \log_2 k / (2c)$.

@@

When $c>$, then $q>4$.
We first show the following equality.
\begin{align}
\left(\frac q 4\right)^r
&=
\left(\frac q 4\right)^{\log_q k}\\
&=
\frac{k}{4^{\log_q k}}\\
&=
\frac{k}{4^{{\log_4 k}/{\log_4 q}}}\\
&=
\frac{k}{k^{1/\log_4 q}}\\
&=
k^{\left(1 - {1}/{\log_4 q}\right)} \\
&=
k^{\left(1 - 1/c\right)}
.
\label{eq:lFU:frac}
\end{align}
Substituting into Eq \eqref{eq:precases}, we have
\begin{align}
    \lF{\star U}^2
    &= \lambda^2B^2 \left(\frac{(q/4)^{r+1}-1}{q/4-1}\right) \\
    \label{eq:lFU:4}
    &= \lambda^2B^2 \left(\frac{(q/4)k^{(1-1/c)}-1}{q/4-1}\right) \\
    &\le \lambda^2B^2 \left(\frac{(q/4)k^{(1-1/c)}}{q/4-1}\right) \\
    &\le 5 \lambda^2B^2 k^{(1-1/c)}
    \label{eq:lFU:5}
\end{align}
    Inequality \eqref{eq:lFU:5} follows because $\tfrac{q/4}{q/4-1} \le 5$ for integer values of $q>4$.
\end{proof}

@@

\section{Experiments}

@@

\begin{table*}
    \centering
    \input{fig/real_world.tex}
    \caption{Experimental results on real world datasets.  For all accuracy measures, larger numbers are better.  The tree loss achieves the best results in all cases.}
    \label{table:results}
\end{table*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plainnat}
\bibliography{paper}
% \begin{abstract}
%   The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
%   both left and right-hand margins. Use 10~point type, with a vertical
%   spacing of 11~points. The \textbf{Abstract} heading must be centered,
%   bold, and in point size 12. Two line spaces precede the
%   Abstract. The Abstract must be limited to one paragraph.
% \end{abstract}

@@

\appendix
\section{Related loss functions}

@@

\end{document}

>>>>

%\newcommand{\fixme}[1]{\noindent{\color{red}\textbf{FIXME:}  {#1}}}
\newcommand{\fixme}[1]{}
\newcommand{\fixmemike}[1]{\noindent{\color{blue}\textbf{FIXME (Mike):}  {#1}}}

@@

\begin{abstract}
    %Multi-class logistic regression has generalization error $O(\sqrt{k/n})$, where $k$ is the number of classes and $n$ the number of data points.

    Multi-class classification problems often have many semantically similar classes.
    For example, 90 of ImageNet's 1000 classes are for different breeds of dog.
    We should expect that these semantically similar classes will have similar parameter vectors,
    but the standard cross entropy loss does not enforce this constraint.
    %These classes are semantically similar to each other,
    %and so we expect their learned parameter vectors to also be similar.
    %Unfortunately, the standard cross entropy loss function does not enforce this constraint.
    %It has generalization error $O(\sqrt{k/n})$, where $k$ is the number of classes and $n$ the number of data points.

@@

%Binary classification is perhaps the oldest and most important problem in statistical learning theory.
%The classic VC theory \citep{vapnik2015uniform} shows that for linear problems in $d$ dimensions and $n$ data points,
%the generalization error decays at a rate of $O(\sqrt{d/n})$.
%Multi-class classification is a natural generalization of the binary case,
%and \citet{natarajan1989learning} generalized the VC theory to show that for a $k$-class problem, the generalization error decays as $O(\sqrt{kd/n})$.
%This bound is tight in the sense that for any given hypothesis class,
%there exists a data distribution whose generalization error will decay as $\Omega(\sqrt{kd/n})$.
%Chapter 29 of \citet{shalev2014understanding} provides a modern overview of these results.

The cross entropy loss is the most widely used loss function for multi-class classification problems,
and stochastic gradient descent (SGD) is the most common algorithm for optimizing this loss.
Standard results show that generalization error decays at a rate of $O(\sqrt{kd/n})$,
where $k$ is the number of class labels, $d$ is the number of feature dimensions, and $n$ is the number of data points.
These results (which we review later) make no assumptions about the underlying distribution of classes.

In this paper,
we assume that the class labels have an underlying metric structure,
and we introduce the tree loss for exploiting this structure.
The tree loss is the first multi-class loss function with provably better generalization error than the cross entropy loss.
We show that if the class label metric has doubling dimension $c$,
then SGD applied to the tree loss will converge at a rate of $O(\sqrt{d \log k/n})$ when $c\le 1$ or $O(\sqrt{k^{1-1/c}d/n})$ when $c\ge 1$.
These improvements are most dramatic for small $c$,
and our empirical results show that many real world classification problems have a label structure with small $c$.

Our paper is organized as follows.
We begin in Section \ref{sec:related} by discussing limitations of related loss functions and how the tree loss will address those limitations.
Section \ref{sec:tree} then formally defines the tree loss.
We emphasize that the tree loss is simply a reparameterization of the standard cross entropy loss,
and so it is easy to implement in common machine learning libraries.
Section \ref{sec:theory} reviews standard results on the convergence of stochastic gradient descent,
and uses those results to prove the convergence bounds for the tree loss.
Finally, Section \ref{sec:experiment} conducts experiments on synthetic, real world image (CIFAR100, ImageNet) and text (Twitter) data.
We show that in practice, the tree loss essentially always outperforms the cross entropy loss.

\section{Related Work}
\label{sec:related}

Previous work on multi-class loss functions have focused on improving either the statistical or computational performance.
Statistical work includes loss functions designed for hierarchichally structured labels \citep{cesa2006incremental,wu2017hierarchical},
loss functions for improving top-$k$ accuracy \citep{lapin2016loss},
or focusing on noisy labels \citep{sukhbaatar2014training,zhang2018generalized}.
The work most similar to ours is the simloss \citep{Kobs2020SimLossCS},
which also makes a metric assumptions about the class label structure.
Our work improves on all of this statistical work by being the first to provide convergence bounds that improve on the bounds of the cross entropy loss.

Other work focuses on improving the speed of multiclass classification.
The hierarchical softmax \citep{morin2005hierarchical} is a particularly well studied variant of the cross entropy loss with many variations \citep[e.g.][]{Peng2017IncrementallyLT,Jiang2017ExplorationOT,Yang2017OptimizeHS,Mohammed2018EffectivenessOH}.
It is easily confused with our tree loss because both loss functions involve a tree structure.
The difference, however, is that the hierarchical softmax focuses on improving runtime performance,
and most variants actually sacrifice statistical performance to do so.
The tree loss, in contrast maintains the runtime performance of the standard cross entropy loss but improves the statistical performance.
Our empirical results in Section \ref{sec:experiment} show that the tree loss has much better statistical performance than the hierarchical softmax.
Other variations of the cross entropy loss include negative sampling, as used in the celebrated word2vec model \citep{mikolov2013distributed}. 
Like with the hierarchical softmax, negative sampling makes no attempt to improve statistical performance.

\ignore{
%Hierarchical classification \citep{cesa2006incremental}
%Hierarchical but without success \citep{wu2017hierarchical}
%
%Focus on top-$k$ accuracy rather than top-1 accuracy \citep{lapin2016loss}
%
%Requires custom training procedure \citep{liu2017easy}
%
%Noisy labels \citep{sukhbaatar2014training,zhang2018generalized}

@@

\end{itemize}
}

@@

\section{The Tree Loss}
\label{sec:tree}

@@

\section{Theoretical Results}
\label{sec:theory}

@@

\begin{lemma}
    \label{lemma:starV}
    We have that for all $V$-trees derived from a $U$-tree,
    \begin{equation}
        \lF{\star V} \le \lF{\star U}.
    \end{equation}
\end{lemma}
\fixme{
\begin{proof}
\end{proof}
}

@@

\begin{lemma}
    \label{lemma:starU}
    Under assumption \ref{ass:B},
    the following bound holds for all $U$-tree structures:
    \begin{equation}
        \lF{\star U} \le 2\sqrt{k}B.
    \end{equation}
\end{lemma}
\begin{proof}
    We have
    \begin{align}
        \lF{\star U}^2 
        &= \sum_{i=1}^k \ltwo{\star\uu_i}^2 \\
        &= \sum_{i=1}^k \ltwo{\star\w_i - \star\w_{\parent{i}}}^2 \\
        &\le \sum_{i=1}^k \left(\ltwo{\star\w_i} + \ltwo{\star\w_{\parent{i}}} \right)^2 \\
        &\le \sum_{i=1}^k (2B)^2 \\
        &\le k (2B)^2
        .
    \end{align}
\end{proof}

@@

\begin{lemma}
    \label{lemma:main}
    When $c\le1$, we have that
    \begin{equation}
        \lF{\star U} \le \tfrac{1}{\sqrt2}\lambda B \sqrt{\log_2 k}.
        \label{eq:c<=1}
    \end{equation}
    When $c>1$, we have that
    \begin{equation}
        \lF{\star U} \le \sqrt{5}\lambda B \sqrt{k^{(1-1/c)}}.
        \label{eq:c>1}
    \end{equation}
\end{lemma}

\fixme{
\begin{proof}
    First, we review some simple properties of the cover tree structure that will be needed in our proof.
    Let $q$ denote the maximum number of children of any node in the cover tree.
    A standard property of cover trees is that $q \le (4^c)$ \citep[Lemma 25]{izbickithesis}.
    Let $Q_i$ denote the number of nodes at level $i$ in the cover tree.
    Then we have that $|Q_i| \le q^i.$
    Finally, let $r$ be the height of the cover tree.
    Without loss of generality, we analyze the bound when the cover tree is perfectly balanced and so $r = \log_q k = \log_2 k / (2c)$.

@@

When $c>$, then $q>4$.
We first show the following equality.
\begin{align}
\left(\frac q 4\right)^r
&=
\left(\frac q 4\right)^{\log_q k}\\
&=
\frac{k}{4^{\log_q k}}\\
&=
\frac{k}{4^{{\log_4 k}/{\log_4 q}}}\\
&=
\frac{k}{k^{1/\log_4 q}}\\
&=
k^{\left(1 - {1}/{\log_4 q}\right)} \\
&=
k^{\left(1 - 1/c\right)}
.
\label{eq:lFU:frac}
\end{align}
Substituting into Eq \eqref{eq:precases}, we have
\begin{align}
    \lF{\star U}^2
    &= \lambda^2B^2 \left(\frac{(q/4)^{r+1}-1}{q/4-1}\right) \\
    \label{eq:lFU:4}
    &= \lambda^2B^2 \left(\frac{(q/4)k^{(1-1/c)}-1}{q/4-1}\right) \\
    &\le \lambda^2B^2 \left(\frac{(q/4)k^{(1-1/c)}}{q/4-1}\right) \\
    &\le 5 \lambda^2B^2 k^{(1-1/c)}
    \label{eq:lFU:5}
\end{align}
    Inequality \eqref{eq:lFU:5} follows because $\tfrac{q/4}{q/4-1} \le 5$ for integer values of $q>4$.
\end{proof}
}

@@

\section{Experiments}
\label{sec:experiment}

@@

\begin{table*}
    \centering
    \input{fig/real_world.tex}
    \caption{Experimental results on real world datasets.  For all performance measures, larger numbers are better.  The tree loss achieves the best results in all cases.}
    \label{table:results}
\end{table*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\bibliographystyle{plainnat}
\bibliography{paper}
% \begin{abstract}
%   The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
%   both left and right-hand margins. Use 10~point type, with a vertical
%   spacing of 11~points. The \textbf{Abstract} heading must be centered,
%   bold, and in point size 12. Two line spaces precede the
%   Abstract. The Abstract must be limited to one paragraph.
% \end{abstract}

@@

\ignore{
\appendix
\section{Related loss functions}

@@

}
\end{document}
